---
title: "JvRad_RNAseqProteomics_Integration"
author: "Raymond Otoo"
date: "2024-01-26"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/")
```

## R Markdown

####Reference using DIABLO from mixOmics with similar data
```{r}
###############################################################################################################################################################################
##      https://github.com/mchimenti/covid_multiomics_mar2022/tree/main

##############################################################################################################################################################################
```


##Download the latest mixomics version from bioconductor
```{r}
# Install BiocManager if not already installed
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::valid()

# install mixOmics
#BiocManager::install("mixOmics")

```
## 1. Load the mixOmics package
```{r}
library(mixOmics)
```


## Read in the finalCounts(table with the genes and counts) for RNA and Proteomics data of proteins and their scores
## 2. Preprocessing of RNAseq count data 
```{r}
# set working directory
setwd("DEG_outputs/")

countdata <- read.table("normalized_counts_significant.txt", header = TRUE, sep = '\t')

#remove the extra column names leaving only the sample names and gene Ids
countdata <- countdata[,!colnames(countdata) %in% c("Chr","Start","End","Strand","Length")]

#transpose rnaseq dataframe while maintaning the column names
countdata_t <- as.data.frame(t(countdata[,-1]))
colnames(countdata_t)<- countdata[,1]

head(countdata_t)
```

## 3.Preprocessing of Proteomics 
```{r}
#This file when received from the core was named as "AllenAR_092723_TMT_protein_results" and other subsequent files will have a similar naming pattern.
#It was then saved as a .csv file and renamed to this.

#read the normalized proteomics data
proteomics_combined <- read.csv("Proteomics_data/TMT_protein_results.csv", header = TRUE)


#remove the extra column names leaving only the sample names and gene Ids
proteomics_combined <- proteomics_combined[,!colnames(proteomics_combined) %in% 
                                             c("MajorityproteinIDs",
                                               "Fastaheaders",
                                               "Score",
                                               "id",
                                               "genesymbol",
                                               "description",
                                               "logFC",
                                               "CI.L",
                                               "CI.R",
                                               "average_intensity",
                                               "t","adj.P.Val",
                                               "sig.PVal",
                                               "sig.FDR",
                                               "B",
                                               "P.Value")]


proteomics_combined
```
## 4. Transpose the proteomics dataframe
```{r}
#transpose proteomics dataframe
proteomics_t <- as.data.frame(t(proteomics_combined), stringsAsFactors = FALSE)
rownames(proteomics_t)<- colnames(proteomics_combined)

#dropped this sample because it was unmatched due to some errors in the mapping of its corresponding sample from the rnaseq. Will be later
#commented out when the error is fixed
#proteomics_t <- proteomics_t[rownames(proteomics_t)!="s204",,drop=F]


#uses the first row as the header
proteomics_new <- proteomics_t
colnames(proteomics_new) <- proteomics_t[1,]

#drops the first row which is now the header
proteomics_new <- proteomics_new[-1,]
proteomics_new
```
## 5. Rename the data so they have the same sample names which will be used to bind them
```{r}
# Extract the first 14 characters from the original column names
short_names <- substr(rownames(countdata_t), 1, 17)

# Remove leading ".." from the short names
short_names <- gsub("^\\.\\.", "", short_names)

# Rename columns using the cleaned short names
rownames(countdata_t) <- short_names
rownames(countdata_t)

new_rownames <- sub("JvRad_Rhipp_","s", rownames(countdata_t))
rownames(countdata_t)<-new_rownames

countdata_t
```
```{r}
short_names
rownames(countdata_t)
```





```{r}
proteomics_new
countdata_t
```
```{r}
#colnames(proteomics_t)
rownames(proteomics_new)
rownames(countdata_t)
```
```{r}
dim(proteomics_new)
dim(countdata_t)
```
```{r}
class(proteomics_new)
class(countdata_t)
```
```{r}
prot_column_types <- sapply(proteomics_new, class)
head(prot_column_types)

rna_column_types <- sapply(countdata_t, class)
head(rna_column_types)
```
```{r}
# reorder the count data
countdata_ordered <- countdata_t[with(countdata_t, order(rownames(countdata_t))), ]

#reorder the proteomics data
proteomics_ordered <- proteomics_new[with(proteomics_new, order(rownames(proteomics_new))), ]


countdata_ordered
proteomics_ordered
```

## Bind the normalized proteomics data to the normalized rnaseq count data
```{r}
library(data.table)
library(dplyr)

# Now use bind_rows
combined_omics <- bind_cols(countdata_ordered, proteomics_ordered)
combined_omics
```
```{r}
rownames(combined_omics)
```

```{r}
#create a list for the reponse variable. Which is the treatment groups
response_list <- c("sham","sham","sham","sham","sham","IR","IR","IR","IR","IR")

response_factor <- factor(response_list)
response_factor

class(response_factor)
```

```{r}

# Convert the proteomics dataframe which is of type "chr" to numeric/double
col_names <- colnames(proteomics_ordered)
row_names <- rownames(proteomics_ordered)

#convert to numeric or double
proteomics_ordered <- apply(proteomics_ordered,2,as.numeric)

#reapply colnames and rownames
colnames(proteomics_ordered) <- col_names
rownames(proteomics_ordered) <- row_names
 
typeof(proteomics_ordered)
```
```{r}
#convert the rnaseq and proteomics dataframes to a datamatix
rna_matrix <- as.matrix(countdata_ordered)
proteomics_matrix <- as.matrix(proteomics_ordered)

#check to be sure they are of the same type
typeof(rna_matrix)
typeof(proteomics_matrix)

#check the dimensions of the matrices
dim(rna_matrix)
dim(proteomics_matrix)

#check to be sure the have the same rownames
rownames(proteomics_matrix)
rownames(rna_matrix)
```


## Clean proteomics data with missing values

```{r}

# Check if there are any missing values in the entire dataframe
any_missing <- any(is.na(proteomics_matrix) | is.nan(proteomics_matrix))

# Print the result
if (any_missing) {
  print("The dataframe contains missing values.")
} else {
  print("The dataframe does not contain missing values.")
}


# Replace missing values with column means
proteomics_matrix[is.na(proteomics_matrix)] <- colMeans(proteomics_matrix, na.rm = TRUE)


# Check if there are any missing values in the entire dataframe
any_missing <- any(is.na(proteomics_matrix) | is.nan(proteomics_matrix))

# Print the result
if (any_missing) {
  print("The dataframe contains missing values.")
} else {
  print("The dataframe does not contain missing values.")
}
```


## Clean RNA data with missing values
```{r}

# Check if there are any missing values in the entire dataframe
any_missing <- any(is.na(rna_matrix) | is.nan(rna_matrix))

# Print the result
if (any_missing) {
  print("The dataframe contains missing values.")
} else {
  print("The dataframe does not contain missing values.")
}


# Replace missing values with column means
rna_matrix[is.na(rna_matrix)] <- colMeans(rna_matrix, na.rm = TRUE)


# Check if there are any missing values in the entire dataframe
any_missing <- any(is.na(rna_matrix) | is.nan(rna_matrix))

# Print the result
if (any_missing) {
  print("The dataframe contains missing values.")
} else {
  print("The dataframe does not contain missing values.")
}
```

```{r}
nzv <- nearZeroVar(rna_matrix)  # No saveMetrics argument

# Get positions of features with zero variance
#features_with_ZeroVar <- which(nzv == TRUE)

# Select corresponding rows from rna_matrix (assuming numerical data)
#zeroVar_features <- rownames(rna_matrix)[features_with_ZeroVar]

```


```{r}
nzv
```


```{r}
#nzv <- rna_matrix %>% 
#  nearZeroVar(saveMetrics = TRUE)

#features_with_ZeroVar <- nzv[nzv$zeroVar == TRUE,]
#features_with_ZeroVar


#select and convert the rownames to a list
#zeroVar_features <- as.list(rownames(a))

```


```{r}
#nzv
#features_with_ZeroVar
```



```{r}
#subset dataframe to exclude zerVar features
#rna_matrix <- rna_matrix[, !colnames(rna_matrix) %in% zeroVar_features, drop=FALSE]

#rna_matrix

```



```{r}
###########################################################  Initial Analysis ############################################################
# Pairwise PLS Comparisons

# set a list of all the X dataframes
X <- list( mRNA = rna_matrix,
            proteomics = proteomics_matrix)
#X[1]

#create the response variable as a factor
response_list <- c("sham","sham","sham","sham","sham","IR","IR","IR","IR","IR")
Y <- factor(response_list)


summary(Y)
# As mentioned in the methods pages, it is strongly advised that prior to using the DIABLO framework that the data be examinsed in a non-integrative context. 
# Here, the correlation between the top 25 features of each dataframe (in a pairwise fashion) are shown in Figure

list.keepX = c(25, 25) # select arbitrary values of features to keep
list.keepY = c(25, 25)

```


```{r}
lapply(X, dim)
```




```{r}
#1. runs the method
results_1 <- block.plsda(X,Y)

#2. Plots the samples
plotIndiv(results_1)

#3. Plots the variables
plotVar(results_1)

```


```{r}
list.keepX <- list(mRNA= c(25,25), proteomics=c(25,25))
list.keepX

results <- block.splsda(X,Y, keepX = list.keepX)
plotIndiv(results)
plotVar(results)
```

```{r}
# Parameter choice
# The design matrix
design <- matrix(0.1, ncol = length(X), nrow = length(X),
                 dimnames = list(names(X), names(X)))

diag(design)<- 0
design
                 
```
```{r}
#Perform Regression analysis with PLS to further understand the correlation between data sets.
#calculate the cross-correlations between components associated to each data set.

# Score was 0.972 indicating that the data sets taken in a pairwise manner are highly correlated. Also, the score suggests that a design with weights ~0.9 could be chosen.

regr1 <- pls(X$mRNA, X$proteomics, ncomp = 1)
cor(regr1$variates$X, regr1$variates$Y)
```

```{r}
## Number of components
# Fit a block.plsda model without variable selection to assess the global performance of the model and choose the number of components to retain.

# run perf() with 10-fold cross validation repeated 10 times for up to five components and with the specified design matrix.


diablo_jvrad <- block.plsda(X,Y, ncomp = 5 , design = design)

perf.diablo_jvrad = perf(diablo_jvrad, validation = 'Mfold', folds = 5, nrepeat = 5)


plot(perf.diablo_jvrad)
```

```{r}
perf.diablo_jvrad$choice.ncomp$WeightedVote
```

```{r}
# choice final ncomp value
ncomp <- perf.diablo_jvrad$choice.ncomp$WeightedVote["Overall.BER","centroids.dist"]

```


```{r}
length(unique(lapply(X, function(x) rownames)))
```

```{r}
# Number of variables to select
test.keepX <- list(mRNA = c(5:9, seq(10, 25, 5)), 
                   proteomics = c(seq(5, 25, 5))) 

tune.diablo.jvrad <- tune.block.splsda(X, Y, ncomp = 2, 
                                      test.keepX = test.keepX, 
                                      design = design, 
                                      validation = 'Mfold', 
                                      folds = 2, 
                                      nrepeat = 1, 
                                      dist = "centroids.dist")

```

```{r}
list.keepX <-tune.diablo.jvrad$choice.keepX
list.keepX
```

```{r}
#final model 
diablo_jvrad_final <- block.splsda(X,Y, ncomp=2,
                                   keepX = list.keepX,
                                   design = design)
```
```{r}
diablo_jvrad_final$design
```
```{r}
#mRNA variables selected on component 1
selectVar(diablo_jvrad_final,block='mRNA', comp = 1)
```

```{r, fig.width=7,fig.height=7,dpi=300}
 png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/plot_diablo.png",width = 10, height = 6, units = "in", res = 600)

plot_diablo <- plotDiablo(diablo_jvrad_final, ncomp = 1)
plot_diablo

dev.off()
```
```{r, fig.width=7,fig.height=7,dpi=300}
 png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/plot_ind.png",width = 10, height = 6, units = "in", res = 600)

plot_ind <- plotIndiv(diablo_jvrad_final, ind.names = FALSE,legend = TRUE,
          title = 'JvRad, DIABLO comp 1 - 2')

plot_ind

dev.off()
```



```{r}
# Plot arrow

plotArrow(diablo_jvrad_final, ind.names = FALSE, legend = TRUE,
          title = 'JvRad, DIABLO comp 1 - 2')
```

```{r, fig.width=7,fig.height=7,dpi=300}

png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/var_plot.png",width = 8, height = 8, units = "in", res = 600)

#variable plots 
plot_var<-plotVar(diablo_jvrad_final, 
        var.names = FALSE, 
        style = 'graphics',
        legend = TRUE,
        pch = c(16,17),
        cex = c(2,2),
        col = c('darkorchid','brown1'),
        title ='JvRad, mRNA vs Proteomics'
)

plot_var

dev.off()
```

```{r, fig.width=7,fig.height=7,dpi=300}

#circos plot
library(ggplot2)

png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/circos_plot.png",width = 10, height = 10, units = "in", res = 600)


plot_circos<- circosPlot(diablo_jvrad_final,
           cutoff = 0.7,
           line= TRUE,
           color.block= c('darkorchid','brown1'),
           color.cor = c('chocolate','grey20'),
           size.labels = 1,
           size.variables = 0.7,
           linkWidth = 1.6,
           showIntraLinks = F,
           size.legend = 0.9,
           )
plot_circos

dev.off()
```

```{r}
#Network
png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/network_plot.png",width = 10, height = 10, units = "in", res = 600)

#plot relevance network
network_plot<-network(diablo_jvrad_final,
        blocks = c(1,2),
        cutoff = 0.4,
        color.node = c('brown1','lightgreen')
        )

network_plot
dev.off()
```

```{r, fig.width=7,fig.height=7,dpi=300}
library(igraph)
my.network = network(diablo_jvrad_final, blocks = c(1,2),
        color.node = c('brown1', 'lightgreen'), cutoff = 0.4)
write_graph(my.network$gR, file = "myNetwork.gml", format = "gml")
```



```{r}
png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/plot_loadings.png",width = 8, height = 4, units = "in", res = 600)

plot_loadings<- plotLoadings(diablo_jvrad_final, comp = 1, contrib = 'max', method = 'median')

plot_loadings
dev.off()

```

```{r}

perf.diablo_jvrad <- perf(diablo_jvrad_final, 
                          validation = 'Mfold', 
                          folds = 2,
                          nrepeat = 1, 
                          dist = 'centroids.dist'
                          )

```




```{r, fig.width=200,fig.height=10,dpi=600}
png(filename = "/Users/raymondotoo/Library/CloudStorage/OneDrive-UniversityofArkansasforMedicalSciences/MultiOmics/Omics Data Integration/01_Transcriptomics_Proteomics/cim_heatmap.png",width = 12, height = 10, units = "in", res = 600)

plot <- cimDiablo(diablo_jvrad_final,
                  legend.position = "topright",
                  comp = c(1,2),
                  size.legend = 1.5,
                  row.names = TRUE,
                  col.names = TRUE,
                  margins = c(8, 18))

plot
dev.off()
```





```{r}
perf.diablo_jvrad$MajorityVote.error.rate
```

```{r}
perf.diablo_jvrad$WeightedVote.error.rate
```

```{r}
auc.diablo.jvrad <- auroc(diablo_jvrad_final,
                          roc.block = 'mRNA',
                          roc.comp = 2,
                          print = FALSE)
```



